Evolving Robot Gaits in Hardware: the HyperNEAT Generative Encoding Vs. Parameter Optimization

### 2011-04-15 paper_ecal/fitnessEvaluation.tex
more more forging
Just about ready to submit

A separate Python tracking server ran on the robot and
interfaced with the Wii remote via bluetooth using the \code{CWiid}
library \citep{cwiid}.  Our fitness-testing code communicated with this
server via a socket connection and requested position updates at the
beginning and end of each run.

@@

\acmFig{linearInterp}{1}{Linear interpolation from the ready position to the first and last commanded gait positions. The position of the robot was measured at
  the beginning and end of each run (red circles) in the same
 ready pose.  Motion was
  interpolated linearly between this stationary pose and the commanded
  gait g(t) for one second at the beginning of the run and two seconds
  at the end, as shown above.}

>>>>

A separate Python tracking server ran on the robot and
interfaced with the Wii remote via bluetooth using the CWiid
library.  Our fitness-testing code communicated with this
server via a socket connection and requested position updates at the
beginning and end of each run.

@@

\acmFig{linearInterp}{1}{Motion was interpolated linearly between a
  stationary pose and the commanded gait $g(t)$ for one second at the
  beginning of each run and two seconds at the end, as shown above.
  The position of the robot was measured at the beginning and end of
  each run (red circles) in the ready pose.}

### 2011-04-15 paper_ecal/hyperNeatMethod.tex
Just about ready to submit

\acmFig{hyperneatExplanation.png}{1}{HyperNEAT produces ANNs from
  CPPNs. ANN weights are specified as a function of the geometric
  coordinates of the source node and the target node for each
  connection. The coordinates of these nodes and a constant bias are
  iteratively passed to the CPPN to determine each connection
  weight. The CPPN has two output values, which specify the weights
  for each connection layer as shown.  Figure from
  \cite{clune2011performance}.}

@@

@@

While measured angles could have been used as input instead of
commanded angles, this may have given HyperNEAT gaits an unfair
advantage by making them closed-loop. Thus, we restricted inputs to
the commanded angles only, which allowed us to produce completely
deterministic gaits using HyperNEAT.

@@

\acmFig{SpiderANN.jpg}{.7}{ANN configuration for HyperNEAT Runs. The
  first two columns of each row of the input layer receive information
  about a single leg (the angles requested in the previous time step for its two joints). The final
  column provides the previously requested angle of the center joint and, to enable periodic movements, a sine and cosine wave. Evolution determines the function
  of the hidden-layer nodes. The nodes in the output layer specify new
  joint angles for each respective joint. The unlabeled nodes in the
  input and output layers are ignored. Figure adapted from
  \cite{clune2011performance}.}\label{hyperneatLayout}

>>>>

\acmFig{hyperneatExplanation.png}{1}{HyperNEAT produces ANNs from
  CPPNs. ANN weights are specified as a function of the geometric
  coordinates of each connection's source and target nodes.
 These coordinates and a constant bias are
  iteratively passed to the CPPN to determine each connection
  weight. The CPPN has two output values, which specify the weights
  for each connection layer as shown.  Figure from
  \cite{clune2011performance}.}

@@

\acmFig{SpiderANN.jpg}{.7}{ANN configuration for HyperNEAT runs. The
  first two columns of each row of the input layer receive information
  about a single leg (the angles requested in the previous time step for its two joints). The final
  column provides the previously requested angle of the center joint and, to enable periodic movements, a sine and cosine wave. Evolution determines the function
  of the hidden-layer nodes. The nodes in the output layer specify new
  joint angles for each respective joint. The unlabeled nodes in the
  input and output layers are ignored. Figure adapted from
  \cite{clune2011performance}.}\label{hyperneatLayout}

@@

%While measured angles could have been used as input instead of
%commanded angles, this may have given HyperNEAT gaits an unfair
%advantage by making them closed-loop. Thus, we restricted inputs to
%the commanded angles only, which allowed us to produce completely
%deterministic gaits using HyperNEAT.

@@

### 2011-04-15 paper_ecal/introduction.tex
Just about ready to submit

\acmFig{robot_whitebg.jpg}{.75}{The quadruped robot. The translucent
  parts were created with 3D-printing technology. Videos of the gaits
  evolved for this robot can be viewed at www.anonymous.url.}

>>>>

\acmFig{robot_whitebg.jpg}{.75}{The quadruped robot for which gaits
  were evolved. The translucent parts were produced by a 3D
  printer. Videos of the gaits can be viewed at http://bit.ly/hLHBYq}

### 2011-04-15 paper_ecal/learningMethods.tex
Just about ready to submit

We implemented two classes of algorithms to produce
the motion-generating function $g(t)$ (described earlier).  The
first method uses a family of parameterized gaits, described in this
section and the next. The second method, HyperNEAT, evolves a non-parametric gait
controlled by neural networks and is described last.

@@

\begin{table}[h!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameters        &                           &       \\
in $\vec{\theta}$ & Description               & Range \\
\hline
\hline
\amp         & Amplitude                 & [0, 400] \\
\hline
$\tau$       & Period                    & [.5, 8] \\
\hline
$m_O$        & Outer-motor multiplier    & [-2, 2] \\
\hline
$m_F$        & Front-motor multiplier    & [-1, 1] \\
\hline
$m_R$        & Right-motor multiplier    & [-1, 1] \\
\hline
\end{tabular}
\caption{The \emph{SineModel5} motion model parameters.}
\tablabel{parameters}
\label{tab:params}
\end{center}
\end{table}

@@

\emph{Uniform random hill climbing}: This method always starts
  with the current best gait, and then chooses the next $\vec{\theta}$ by
  randomly choosing one parameter to adjust and replacing it with a
  new value chosen with uniform probability in the allowable range for
  that parameter. This new point is evaluated, and if it results in a
  longer distance walked than the previous best gait, it is saved as
  the new best gait. The process is then repeated.

>>>>

%We implemented two classes of algorithms to produce the
%motion-generating function $g(t)$ (described earlier).  The first
%method uses a family of parameterized gaits, described in this section
%and the next. The second method, HyperNEAT, evolves a non-parametric
%gait controlled by neural networks and is described last.

We now describe the gait-generating algorithm classes.

@@

\begin{table}[b]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameters        &                           &       \\
in $\vec{\theta}$ & Description               & Range \\
\hline
\hline
\amp         & Amplitude                 & [0, 400] \\
\hline
$\tau$       & Period                    & [.5, 8] \\
\hline
$m_O$        & Outer-motor multiplier    & [-2, 2] \\
\hline
$m_F$        & Front-motor multiplier    & [-1, 1] \\
\hline
$m_R$        & Right-motor multiplier    & [-1, 1] \\
\hline
\end{tabular}
\caption{The \emph{SineModel5} motion model parameters.}
\tablabel{parameters}
\label{tab:params}
\end{center}
\end{table}

@@

\emph{Uniform random hill climbing}: This method repeatedly starts
  with the current best gait and then selects the next $\vec{\theta}$ by
  randomly choosing one parameter to adjust and replacing it with a
  new value chosen with uniform probability in the allowable range for
  that parameter. This new point is evaluated, and if it results in a
  longer distance walked than the previous best gait, it is saved as
  the new best gait.

### 2011-04-15 paper_ecal/platformDetails.tex
more more forging
Just about ready to submit

All of the computation for gait learning, fitness evaluation, and
robot control was performed on the compact, on-board CompuLab Fit-PC2,
running Ubuntu Linux 10.10. 
The most computationally expensive portion of code was HyperNEAT, but even
this took less than one second to run on the on-board computer. Thus,
we chose not to offload any computation.
All gait generation, learning, and fitness
evaluation code, except HyperNEAT, was written in Python and is
available on our website \citep{quadratotWeb}.  HyperNEAT is written in
C++.  We controlled the servos with the Pydynamixel library
\citep{pydynamixel}, sending commanded positions at 40Hz.  The robot
connected to a wireless network on boot, which enabled us to control
it via SSH.

>>>>

All of the computation for gait learning, fitness evaluation, and
robot control was performed on the compact, on-board CompuLab Fit-PC2,
running Ubuntu Linux 10.10. 
The most computationally expensive portion of code was HyperNEAT, but even
this took less than one second to run on the on-board computer. Thus,
we chose not to offload any computation.
All gait generation, learning, and fitness
evaluation code, except HyperNEAT, was written in Python and is
available on our website (http://bit.ly/hLHBYq).  HyperNEAT is written in
C++.  We controlled the servos with the Pydynamixel library, sending commanded positions at 40Hz.  The robot
connected to a wireless network on boot, which enabled us to control
it via SSH.

### 2011-04-15 paper_ecal/relatedWork.tex
Just about ready to submit

Various machine learning techniques have proved to be effective at
generating gaits for legged robots. Kohl and Stone presented a policy
gradient reinforcement learning approach for generating a fast walk on
legged robots \citep{kohl}, which we implemented for comparison. Others
have evolved gaits for legged robots, producing competitive
results \citep{chernova2005evolutionary, hornby2005autonomous, zykov,
  clune2009evolving, clune2011performance, clune2009hybrid,
  clune2009sensitivity, tellez2006evolving, valsalam2008modular}. In
fact, an evolved gait was used in the first commercially-available
version of Sony's AIBO robot~\citep{hornby2005autonomous}. Except for
work with HyperNEAT \citep{clune2009evolving, clune2011performance,
  clune2009hybrid, clune2009sensitivity}, the previous evolutionary
approaches have helped evolution exploit the regularity of the problem
by manually decomposing the task. Experimenters have
to choose which legs should be coordinated, or otherwise facilitate
the coordination of motion. Part of the motivation of this paper
is to see if the space of regularities can be explored more systematically,
which has previously performed well \citep{valsalam2008modular}, and to
compare such an approach to HyperNEAT.

>>>>

Various machine learning techniques have proved to be effective at
generating gaits for legged robots. Kohl and Stone presented a policy
gradient reinforcement learning approach for generating a fast walk on
legged robots \citep{kohl}, which we implemented for
comparison. Others have evolved gaits for legged robots, producing
competitive results \citep{chernova2005evolutionary,
  hornby2005autonomous, zykov, clune2009evolving,
  clune2011performance, clune2009hybrid, clune2009sensitivity,
  tellez2006evolving, valsalam2008modular}. In fact, an evolved gait
was used in the first commercially-available version of Sony's AIBO
robot~\citep{hornby2005autonomous}. Except for work with HyperNEAT
\citep{clune2009evolving, clune2011performance, clune2009hybrid,
  clune2009sensitivity}, the previous evolutionary approaches have
helped evolution exploit the regularity of the problem by manually
decomposing the task. Experimenters have to choose which legs should
be coordinated, or otherwise facilitate the coordination of
motion. Part of the motivation of this paper is to compare the
regularities produced by HyperNEAT to those generated by a more
systematic exploration of regularities via a parameterized model.
